### 2022-03-17 draft.tex
Deploying to gh-pages from @ gottacatchenall/ms_neutral_landscapes_jl@64381af074f3e728fd9ea4ee40d6a3f206792be7 🚀

Neutral landscapes are increasingly used in ecological and evolutionary
studies to provide a null expectation spatial variation of a given
measurement. Originally developed to simulate the spatially
autocorrelated data (Gardner \emph{et al.} 1987; Milne 1992), the have
seen use in a wide range of disciplines: from landscape genetics
(Storfer \emph{et al.} 2007), to landscape and spatial ecology (Tinker
\emph{et al.} 2004; Remmel \& Fortin 2013), and biogeography (Albert
\emph{et al.} 2017).

We present \texttt{NeutralLandscapes.jl}, a package in \texttt{Julia}
for neutral landscapes. The two primary packages used to simulate
neutral landscapes are \texttt{NLMR} in (the \texttt{R} language)
(Sciaini \emph{et al.} 2018) and \texttt{NLMpy} (in Python; Etherington
\emph{et al.} 2015). Here we demonstrate that
\texttt{NeutralLandscapes.jl}, depending on the method, is orders of
magnitude faster than previous neutral landscape packages.

As biodiversity science becomes increasingly concerned with temporal
change and its consequences, its clear there is a gap in methodology in
generating neutral landscapes that change over time. In addition we
present a novel method for generating landscape change with prescribed
levels of spatial and temporal autocorrelation, which is implemented in
\texttt{NeutralLandscapes.jl}

@@

@@

quebec = SimpleSDMPredictor(WorldClim, BioClim; left=-90., right=-50., top=75., bottom=40.)
qcmask = fill(true, size(quebec))
qcmask[findall(isnothing, quebec.grid)] .= false

@@

@@

\hypertarget{generating-dynamic-neutral-landscapes}{%
\section{Generating dynamic neutral
landscapes}\label{generating-dynamic-neutral-landscapes}}

\hypertarget{temporally-autocorrelation}{%
\subsubsection{Temporally
autocorrelation}\label{temporally-autocorrelation}}

\(r\): rate, \(v\): variability, \(U\) matrix of draws from standard
\(\text{Normal}(0,1)\).

Here \(v\) replects the amount of temporal autocorrelation.

\(f_{T}(M_{ij}) = r + vU_{ij}\)

\hypertarget{spatial-autocorrelation}{%
\subsubsection{Spatial autocorrelation}\label{spatial-autocorrelation}}

\(r\): rate, \(v\): variability, \([Z(\delta)]_{ij}\): the \((i,j)\)
entry of the zscore of the \(\delta\) matrix

\hypertarget{spatiotemporal-autocorrelation}{%
\subsubsection{Spatiotemporal
autocorrelation}\label{spatiotemporal-autocorrelation}}

@@

>>>>

\begin{itemize}
\tightlist
\item
  what are neutral landsacpes
\end{itemize}

Neutral landscapes are increasingly used in ecological and evolutionary
studies to provide a null expectation spatial variation of some
measurement.

Originally based around methods for simulating spatially autocorrelated
data (Gardner \emph{et al.} 1987; Milne 1992).

whawt is spaital autocorrelation

they have seen use in a wide range of fields in ecology and evolution:
from landscape genetics (Storfer \emph{et al.} 2007), to landscape and
spatial ecology (Tinker \emph{et al.} 2004; Remmel \& Fortin 2013), and
biogeography (Albert \emph{et al.} 2017).

The two most popular libraries used to simulate neutral landscapes are
\texttt{NLMR} in (the \texttt{R} language) (Sciaini \emph{et al.} 2018)
and \texttt{NLMpy} (in Python; Etherington \emph{et al.} 2015). Here we
present \texttt{NeutralLandscapes.jl}, a package for neutral landscape
simulation in the \texttt{Julia} language. So, why another package? Here
we demonstrate that \texttt{NeutralLandscapes.jl} is orders of magnitude
faster than previous neutral landscape packages. In addition,
\texttt{NeutralLandscapes.jl} implements several novel methods for
simulating environmental change with temporal variation.

As biodiversity science becomes increasingly concerned with temporal
change and its consequences, its clear there is a gap in methodology in
generating neutral landscapes that change over time. Our model allows
users to simulate time-series of any \texttt{NeutralLandscape} layer,
and which produce an arbitrary distribution of change across every
spatial cell, with provided levels of spatial and temporal
autocorrelation.

@@

Further, \texttt{NL.jl} provides methods for interacting with other
\texttt{julia} packages, and functions for rescaling

@@

quebec = SimpleSDMPredictor(WorldClim, BioClim; left=-90., right=-50., top=75., bottom=40.)
qcmask = fill(true, size(quebec))                   # ------ TODO
qcmask[findall(isnothing, quebec.grid)] .= false    # should both of these lines be possible only using mask?

@@

Figure: Real temp (left) / Rescaled NL (right) , same unit bar

\hypertarget{generating-dynamic-neutral-landscapes}{%
\subsection{Generating dynamic neutral
landscapes}\label{generating-dynamic-neutral-landscapes}}

Two types of temporal change: (1) null change, where the is random
variation in each cell across time but the mean value across all cells
stays constant (with some variation around this constant due to
randomness in change generation)

\hypertarget{null}{%
\subsubsection{Null}\label{null}}

Take an arbitrary distribution (from \texttt{Distributions.jl}) and set
its mean value to 0, and apply draws from that distribution to each cell
at each timestep.

Can take an arbitrary distribution of values and set its expected-value
to be the primary input into a change model---the mean amount of change
at each timestep. This can also be parameterized to a be a variable list
of mean change at each corresponding timestep.

\hypertarget{temporally-autocorrelation}{%
\paragraph{Temporally
autocorrelation}\label{temporally-autocorrelation}}

We generate temporally autocorrelated change using the method. We take
an arbitrary distribution \(A\)

\(r\): rate, \(v\): variability, \(U\) matrix of draws from standard
\(\text{Normal}(0,1)\). Here \(v\) replects the amount of temporal
autocorrelation.

The value of a given cell \((i,j)\) with value \(M_{ij}\)
\[M_{ij}(t+1) = f_{T}(M_{ij}(t)) = r + vA_{ij}\]

\hypertarget{spatial-autocorrelation}{%
\paragraph{Spatial autocorrelation}\label{spatial-autocorrelation}}

\(r\): rate, \(v\): variability, \([Z(\delta)]_{ij}\): the \((i,j)\)
entry of the z-score of the \(\delta\) matrix

Z-score is arbitrary and can be replaced with any dist.

\hypertarget{spatiotemporal-autocorrelation}{%
\paragraph{Spatiotemporal
autocorrelation}\label{spatiotemporal-autocorrelation}}

Finally, to implement change this is both spatially and temporally
autocorrelated

@@

How many lines of code, and what language is that code in for each pkg?
\texttt{NLMR} contains 893 lines of R, and 376+51 lines of C++.
\texttt{nlmpy} contains 386 lines of python. Julia contains 664 lines of
(non-test) \texttt{julia}. Note these numbers refer only to lines of
code and not comments.

\hypertarget{example-fitting-a-neutral-landscape-to-an-empirical-spatial-dataset}{%
\section{Example: fitting a neutral landscape to an empirical spatial
dataset}\label{example-fitting-a-neutral-landscape-to-an-empirical-spatial-dataset}}

Here we use approximate bayesian computation to estimate the parameter
of autocorrelation H for an empirical raster of temperature data.

Why? What if we are interested in differentiating the processes that
occur in this \emph{real} landscape versus landscapes with \emph{similar
statistical properties}.

We take a raster of mean temp around the st lawrence lowlands in QC, and
use ABC to estimate the value of H under the midpoint-displacement
model.

We use the variogram as the loss function

\begin{figure}
\hypertarget{fig:post}{%
\centering
\includegraphics{./figures/posterior.png}
\caption{todo}\label{fig:post}
}
\end{figure}

@@

Why is it good that we've made this a faster thing to do? Why are models
of temporal change necessary? What can simulation do for spatial ecology
more generally?

What are questions we can address with NL.jl that was not possible
before?
